---
author: "Jieyi Zhu(Jz6282)"
date: "BIO382K Project1"
fig.width: 4
output:
  word_document: default
  html_notebook: default
fig.height: 3
---

### 1. IBM population simulation description
+ There are two color phenotypes: Red and Orange. 
+ This model will consist of a collection of reproducing agents.
+ Reproduction and mortality will be stochastic functions, with reproductive rate depending on the total number of agents

### 2.Parameter descriptions and definitions

The model is 
$$
N_{t+1} - N_t = Birth_t - Death_t
$$
where $Birth_t \sim \mathrm{Poisson}(\lambda)$, $\lambda = N_t e^{r_0\left(1 - \frac{N_t}{K}\right)}$ and $Death_t \sim \mathrm{Binom}(N_t, \delta)$. These function give random births and deaths that occur over one update step.

+ $N_t$ : The current population size
+ $N_{t+1}$ : The population size in next time interval
+ $r_0$ : The maximum possible growth rate 
+ $K$ : The population size where each mother will produce one offspring. 
+ $Birth_t$ : The total number of offspring
+ $Death_t$ : The total number of death.
+ $\delta$ : The mortality rate in each time interval


```{r}
Colors = c("orange","red") #population phenotypes
t_max = 100 #  maximum time duration of simulation
r0 = log(2) # maxium births is 2 per mother
K = 100 # population size where births is 1 per monther
delta = 0.1 # natural mortality
N0 = 5*length(Colors) # initial number of individuals
```

### 3. Initializing the model

+ Initialize the model with a function which conatins the ID, Colors, and age
+ Store the population size of each colore phenotype into a data frame.

```{r}
# give the starting attributes (id, color, age)
make.inds <- function(id=NaN, color=NaN, age=NaN){
  inds <- data.frame(id = id, color = color, age = 0) 
  inds
}

# initial population size of each color phenotype
inds <- make.inds(id=1:N0, color=as.factor(array(Colors, dim=N0))) 

# store the population sizes of each color phenotype
N <- data.frame(NaN, dim=c(t_max + 1, length(Colors)))
colnames(N) <- Colors
N[1,] <- summary(inds$color) 

```

### 4.Simulation
+ looping the individual in the simulation

```{r}
set.seed(12345)
for (t in 1:t_max + 1) # loop over time
{
  M =  length(inds$age)# population size
  lambda = exp(r0 * (1 - M / K)) 
  for (i in 1:M) # loop over individuals
  {
    j = rbinom(1, 1, delta) # will return a zero or one
    if (j == 1) inds$age[i] = -1 # if j equals 1, mark as dead
    else inds$age[i] = inds$age[i] + 1 # if j not equals to 1, the age will incrase by 1
    n_offsp = rpois(1, lambda) # the number of offspring per mother
    if(n_offsp>0){
      offspring <- make.inds(id=seq(max(inds$id)+1, length.out=sum(n_offsp)))
      offspring$color <- inds[i,"color"]
      inds = rbind(inds, offspring) # append the offrping to end of inds
    }
  }
  inds= inds[inds$age != -1,] # exclude dead individuals
  N[t,] = summary(inds$color) # put the population size of each color in N data frame
}

pop_data = data.frame(t = 0:t_max, N) # combine the population size and tiem interval
pop_data$total = pop_data$orange + pop_data$red # Get the total population size in each time interval


```

### 5. Plots
+ Plot the population size in each color and plot the total population in each time interval
+ Plot the Age distribution

```{r}
library(ggplot2)
# The population size plot
cols <- c("Orange Pop."="orange","Red Pop."="red","Total Pop."= "steelblue")
p_pop <- ggplot(pop_data, aes(x = t))
p_pop <- p_pop+ geom_line(aes(y = orange, col="Orange Pop."), lwd = 1 ) +
        geom_line(aes(y = red, color = "Red Pop."),lwd = 1) +
        geom_line(aes(y = total, color = "Total Pop."),lwd = 1)+ ylab("Population Size")

p_pop  <- p_pop +scale_colour_manual(values=cols)+ theme_bw() + ggtitle("The Population Simulation") +
        theme(plot.title = element_text(hjust = 0.5, size = 15)) 
p_pop
 
# Age distribution
p_age <- ggplot(inds, aes(x = age)) + geom_histogram(color='black', bins=50)
p_age <- p_age + ggtitle("Age Distribution") + theme_bw()+theme(plot.title = element_text(hjust = 0.5, size=15))
p_age
```

